Laurentia's continued motion into the Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation

### 2022-07-20 Manuscript/Manuscript.tex
Updates from Overleaf

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{amsmath}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

@@

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of Laurentia's plate speed is the onset of the ca. 1090-980 Ma Grenvillian Orogeny along the eastern margin. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum deposition age has been constrained to be ca. 992 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along 5 stratigraphic sections in northern Michigan, USA. High-resolution thermal demagnetization experiments resolve detrital magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, providing constraints for the duration of the Keweenawan superchron. Inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the development of supercontinent Rodinia.
% rather than recording a significant crustal shortening.
\end{abstract}

@@

\begin{keypoints}
\item Neoproterozoic
\item Laurentia  
\item Paleogeography
\item apparent polar wander path
\item Inclination shallowing
\item Grenville
\item Keweenawan superchron
\item Red bed
\end{keypoints}

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provides crucial constraints on global paleogeography in the Mesoproterozoic and has become a central record for reconstructing the development of ancient supercontinent Rodinia \cite{Swanson-Hysell2019a, Swanson-Hysell2021c}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a quiescence period that could have lasted as much as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. As a result, uncertainties exist in reconstructing Laurentia and its conjugate continents' configuration in the late Mesoproterozoic to early Neoproterozoic.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing location of the Grenville Front relative to the study area. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, and major post-rift thrust faults. Jacobsville stratigraphic sections along which paleomagnetic samples were collected are shown as green diamond symbols.}
\label{fig:Geologic_map}
\end{figure*}

Although igneous activities are lacking, sedimentary rocks formed near the end of the Midcontinent Rift magmatism and afterwards, and can provide additional paleomagnetic data. In particular, hematite-bearing sedimentary rocks (i.e. red beds) can provide insights into the ancient geomagnetic field and the paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provide crucial constraints for Laurentia's latitude, orientation, and plate motion in the late Mesoproterozoic \cite{Henry1977a}.

In addition to the Oronto Group sediments, the overlying post-rift Jacobsville Formation which were deformed during the inversion of the Midcontinent Rift also contains abundant red beds which bear the opportunity to further extend the reconstruction of the apparent polar wander path of Laurentia toward more recent times \cite{Hamblin1958a, Hodgin2022a}. U-Pb detrital zircon ages and U-Pb calcite ages from late-kinematic veins constrain the maximum depositional age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Laurentia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records as they acquire magnetic remanence not through cooling from above the magnetic ordering temperatures of magnetic minerals as igneous rocks do, but acquire detrital remanence as magnetic particles settle from the water column during deposition, and chemical remanence as magnetic minerals precipitate and grow within pore spaces of host sedimentary rocks syn- to post-deposition (i.e. pigmentary hematite). It has been found that the formation of the pigmentary hematite in red beds can significantly post-date the timing of sediment deposition, resulting in the chemical remanent magnetization being secondary, superimposed upon the primary remanent magnetization carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974, Roy1978} adopted a method first applied by \citeA{Collinson1965a} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Paired acid etching demagnetization and thermal demagnetization experiments show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Bilardello2010a, Tauxe1980a}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures \cite{Swanson-Hysell2019b}. In particular, \cite{Swanson-Hysell2019b} conducted a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation to show that via thermal demagnetization one can isolate the primary magnetization (which are reoriented in random directions in the clasts) from the secondary component carried by the co-occurring authigenic hematite (which yield similar directions amongst clasts).

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Taux2004b, Bilardello2016b}. Due to the rotation of detrital magnetic grains toward bedding-parallel caused by authigenic processes such as compaction during lithification, inclinations recorded by sediments often have shallower angles than what was during the time of deposition. To correct for such bias, \citeA{Tauxe2004b} developed a statistical model based on the geocentric axial dipole (GAD) hypothesis where the amount of inclination shallowing in a specific sediment package is evaluated based on the shape of the distribution of the sedimentary paleomagnetic directions is compared to that predicted by a paleosecular variation model developed for the past 5 Myr. \citeA{Pierce2022a} showed that the paleosecular variation model of \citeA{Tauxe2004b} is also suitable for application to sedimentary rocks formed during the Midcontinent Rift. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the commonly used circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation. High-resolution thermal demagnetization data on centimeter-scale siltstone intraclasts that were eroded by fluvial processes and redeposited at two locations constrain the timing of hematite acquisition by revealing a primary component that formed prior to the erosion of the clasts within the depositional environment and a secondary component that formed following their redeposition. Then we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia and its association with Baltica during the development of the supercontinent Rodinia.

% preservation of the Midcontinent Rift sedimentary rocks (Figure \ref{fig:Geologic_map}), complex magnetic remanence components that often form during significantly different times within the rocks and post-deposition modification of the remanence magnetization directions are common to sedimentary rocks, presenting challenges to the interpretation of their paleomagnetic records \cite{Beck2003a, Butler1992a, Vandervoo2012a}. In contrast to igneous rocks which acquire thermal remanent magnetizations through cooling of iron oxides, red beds often record mixtures of magnetic remanence components during primary hematite grain deposition and secondary hematite precipitation and growth that are often not coeval \cite{Tauxe1980a, Swanson-Hysell2019b}. Detrital hematite grains can record magnetic remanence at the time of deposition, whose age can be constrained by the maximum deposition age of the host sedimentary rock. The secondary source of hematite is the grains that precipitate and grow in situ within pore space of host sedimentary rocks that have been deposited (i.e. pigmentary hematite). These secondary hematite could form significantly after deposition of the sedimentary rocks and can keep growing through a protracted period of time. To isolate the primary magnetization component from the secondary one within red beds is crucial for accurately reconstructing paleomagnetic records at the time of red beds deposition. Another challenge to interpreting sedimentary paleomagnetic data is the problem of inclination shallowing \cite{King1955a}. Grain rotation toward being bedding plane parallel during processes associated with deposition and lithification can result in recorded inclinations being shallower than the local magnetic inclination. Such deviation may vary depending on the location where the sedimentary rock formed and can result in various degrees of uncertainties of paleomagnetic reconstructions.

@@

% Previous paleomagnetic data from the Midcontinent Rift rocks that are paired with high-precision geochronology data has led to the development of the Keweenawan Track, showing rapid plate motion (latitudinal motion of $\sim$ cm/yr) from high latitudes between ca. 1105 Ma and ca. 1092 Ma, followed by a decrease in plate motion (latitudinal motion of $\sim$) between ca. 1092 Ma and ca. 1080 Ma when Laurentia moved to the equator and ceased its rapid motion. In particular, paleomagnetic poles developed from the ca. 1080 Ma Nonesuch Formation and the ca. 1050 Ma upper Freda Formation of the Oronto Group sediments indicate that Laurentia's plate speed significantly reduced in the late Mesoproterozoic, resulting in little change in its paleocontinental configuration for tens of millions of years.

@@

% \cite{Roy1978a} has done that but without the effort of distinguishing the detrital remanence and the chemical rmanence.
% In addition, constraining the pole position of the

% and shed new lights on reconciling the connection between the Keweenawan Track and the Grenville Loop paleomagnetic apparent polar wander paths.

@@

% In this study, we use fine-grained to silt-sized, hematite rich red beds from xxx sections along the Keweenaw Peninsula, Michigan with the goal of obtaining high-quality paleomagnetic pole from the jacobsville formation that is corrected for inclination shallowing problem and to investigate the connection between the Keweenawan Track and the Grenville Loop of the Laurentia APWP during the late Mesoproterozoic to the Neoproterozoic.

@@

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through Laurentia craton (Figure \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1080 Ma to 1050 Ma Oronto Group sediments \cite{Swanson-Hysell2019a, Fuentes2022a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \cite{Cannon1992a}). Subsequently, the rift was inverted as the Grenvillian Orogeny commenced on the eastern margin of Laurentia (Figure \ref{fig:Geologic_map}; \citeA{Cannon1994a}), resulting in the propagation of far-field compressional stress deep into the interior of Laurentia. During the inversion, Paleoproterozoic and Archean crusts were uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \cite{Cannon1993a}). The post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a, Kalliokoski1982a}). The erosional contact and the development of paleosol between the Jacobsville and the underlying basement rocks is consistent with the interpretation that there was a prolonged period of sedimentation quiescence and net erosion between the formation of the Oronto Group sediments and the Jacobsville Formation. The overall extent of the Jacobsville Formation has been mapped along the lake shore of the Keweenaw Peninsula and is interpreted to extend eastward to Sault Ste Marie, Canada (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a}).

% The Jacobsville Sandstone is unconformably overlain by the Cambrian Munising Formation (Hamblin, 1958).

As the far-field stress of the Grenvillian Orogeny continued to reach the interior of Laurentia, it thrusted and folded the Midcontinent Rift volcanics over the Jacobsville strata along the major Keweenaw fault which extends $\sim$250 km from the tip of the Keweenaw Peninsula in northern Michigan southeastward to a termination in northeastern Wisconsin (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1996a, Cannon2001a, DeGraff2022a}). As a result, rift volcanic clasts where eroded into the Jacobsville basin at the same time when sediments from the older basement rocks along with distal input from the Grenville Province were deposited within the Jacobsville back-bulge basin system that in general wedges toward the northwest (Figure \ref{fig:Geologic_map}; \cite{Hedgman2014a, Hodgin2022a}. Immediately near the Keweenaw fault in the footwall, folding has also been recorded by strata of the upper Jacobsville Formation in contact with the volcanics \cite{Cannon1993a}. One famous remanence of that late-stage deformation is the Natural Wall ravine where sub-vertical to overturned sandstone beds are preserved (Figure \ref{fig:Geologic_map}; \cite{Hamblin1958a}). On shore of Bete Grise Bay near the northeastern edge of the Keweenaw Peninsula, the Keweenaw fault is interpreted to have accommodated compressional stresses by forming a series of segmented en echelon thrusts \cite{Tyrrell2019a}. The timing of such en echelon faulting could be coeval with the deposition of the upper Jacobsville Formation, leading to the nonconformable onlapping of the Jacobsville sediments over the volcanics in addition to the thrust faulting relationship between the two units. Late-kinematic calcite that precipitated within the brecciated Keweenaw fault zone and youngest detrital zircons of the Jacobsville Formation constrain the timing of the final inversion of the Keweenaw Fault to be during the Rigolet phase of the Grenvillian Orogeny \cite{Hodgin2022a}.

\subsection{General stratigraphy}

@@

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose sandstones, conglomerates, siltstones, and shales, completely devoid of lava flows or cross-cutting dikes. It is in general more quartz-rich than the Oronto Group sediments. Abundant near the bottom section in the southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Keweenaw, Marenisco, and Pelton Creek thrust faults is interpreted to represent syn-depositional development of local relief \cite{Kalliokoski1982a, Hedgman1992a}).

Near the Keweenaw Fault (Fig. \ref{fig:Geologic_map}), the Jacobsville Formation consists dominantly of quartz-rich, cross-bedded, medium-grained sandstone, locally abundant clast-supported conglomerate, with interbeds of hematite-bearing micaceous siltstone (Fig. \ref{fig:strat_column}). In this region, the conglomerate contains abundant volcanic clasts likely derived from uplifted Midcontinent Rift volcanics \cite{Irving1885a, Brojanigo1984a}. Occurrence of red siltstone beds have been reported in the Jacobsville Formation, particularly in the upper sections near Keweenaw Fault \cite<e.g.>{Hamblin1958a, Irving1885a}. In the field, they typically appear brick-red to dark-red, consisting of find-grained arenite to arkosic sandstone and siltstone sometimes with minor grey reduction spots (Fig. \ref{fig:Field_photo}). Rounded detrital mica grains up to $\sim$5mm in size can be found commonly within very fine, fissile siltstone and shaly strata (Figure \ref{fig:Field_photo}). The fact that in the Jacobsville has conglomerate and sandstone and siltstone facies in them supports sediment input from the eroded Oronto Group. This indicate the older boundary of the Jacobsville deposition might be closer to the Rigolet Phase than the Ottawan phase and the duration of Jacobsville sedimentation may be within a relatively constrained period of time.

tyrrell 2019 fig. 10 recorded onlapping of the Jacobsville Formation on to the Portage Lake Volcanics which appears to have been faulted prior to the deposition of Jacobsville near Bete Grise Bay.

% Hematite is one of the most stable form of iron oxides under ambient conditions, and magnetic remanence carried by the grains within red beds can be preserved for billions of years \cite{Butler1992a, Jiang2022a}. Previously, paleomagnetic records of the Jacobsville Formation have been investigated by \citeA{Roy1978a} who sampled a range of lithofacies of the Jacobsville Formation and obtained that a primary remanence component can be isolated via magnetic cleaning. In this study, we target the find-grained sandstone to siltstone facies in the upper Jacobsville Formation and segregate the primary and secondary remanent magnetizations using high-resolution thermal demagnetization method \cite{Swanson-Hysell2019b}. We update the paleomagnetic records of the Jacobsville Formation with additional evaluation for the degree inclination shallowing that \citeA{Roy1978a} did not take into account.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=0.9\textwidth]{strat_column.png}
\caption{}
\label{fig:strat_column}
\end{figure*}

Keweenaw Fault motion post dating Jacobsville in the footwall
Youngest DZ from jacobsvill eis about 993 Ma (sandstone creek, about a few hundred meters from the fault). 
directly above pmag sampling at Agate falls grain B geochron date is (z2, s226, about 1000 Ma grain)

>>>>

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{sidecap}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

@@

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of Laurentia's plate speed is the onset of the ca. 1090-980 Ma Grenvillian Orogeny along the eastern margin. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum depositional age has been constrained to be ca. 992 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along five stratigraphic sections in northern Michigan, USA. High-resolution thermal demagnetization experiments resolve detrital magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, providing constraints for the maximum duration of the Keweenawan superchron. An inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the development of the supercontinent Rodinia.
% rather than recording a significant crustal shortening.
\end{abstract}

@@

\section*{Keywords}
Neoproterozoic\\
Laurentia  \\
Paleogeography\\
\

apparent polar wander path\\
Inclination shallowing\\
Grenville\\
Keweenawan superchron\\
Red bed\\

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provide crucial constraints on global paleogeography in the Mesoproterozoic \cite{Swanson-Hysell2019a}. As a result, the Keweenawan Track has become a central record for reconstructing the development of the ancient supercontinent Rodinia \cite{2021a}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a period of quiescence that could have lasted for as long as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a}. Due to this magmatic gap, Laurentia and its conjugate continents' paleogeographic configuration in the late Mesoproterozoic to mid-Neoproterozoic has remained largely uncertain \cite{Swanson-Hysell2021c}.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing location of the Grenville Front relative to the study area. The inset green box shows the extent of panel B. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, other older Proterozoic rocks, and major post-rift thrust faults. Jacobsville stratigraphic sections along which paleomagnetic samples were collected in this study are shown as green diamond symbols.}
\label{fig:Geologic_map}
\end{figure*}

When igneous rock records are lacking, paleomagnetic data recorded by sedimentary rocks (e.g. red beds) can provide useful insights into the ancient geomagnetic field and record paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provide crucial constraints for Laurentia's latitude, orientation, and plate motion in the late Mesoproterozoic \cite{Henry1977a, Swanson-Hysell2019a, Swanson-Hysell2019b}.

On top of the Oronto Group sediments overlies the post-rift Jacobsville Formation which was partly deformed during the inversion of the Midcontinent Rift (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). The Jacobsville Formation also contains abundant red beds which offer the opportunity to further extend the reconstruction of the apparent polar wander path of Laurentia toward more recent times \cite{Hamblin1958a, Roy1978a}. U-Pb detrital zircon ages and U-Pb calcite ages from late-kinematic veins constrain the maximum depositional age of the Jacobsville Formation to be ca. 992 Ma \cite{Hodgin2022a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Laurentia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records. Primary detrital remanence acquired during settling of magnetic particles in the water column can be masked as secondary magnetic minerals (e.g. pigmentary hematite) that precipitate and grow within pore spaces of host sedimentary rocks syn- to post-deposition. In red beds, it has been found that formation of the pigmentary hematite can significantly post-date the timing of sediment deposition, resulting in the chemical remanent magnetization being secondary, superimposed upon the primary remanent magnetization carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974a, Roy1978a} adopted a method first applied by \citeA{Collinson1965c} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Studies applying paired acid etching demagnetization and thermal demagnetization show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Tauxe1980a, Bilardello2010c}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures \cite{Swanson-Hysell2019b}. In particular, a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation shows that via thermal demagnetization one can isolate detrital magnetization vectors (which are reoriented in random directions in the clasts) from the secondary components carried by co-occurring authigenic hematite (which yield similar directions amongst clasts) \cite{Swanson-Hysell2019b}.

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Tauxe2004b, Bilardello2016b}, which is necessary due to the rotation of detrital magnetic grains to be parallel to bedding caused by authigenic processes such as compaction during lithification. As a result, inclinations recorded by sediments often have shallower angles than the local field inclinations during the time of deposition. To estimate the amount of inclination shallowing in a given package of sedimentary rocks, \citeA{Tauxe2004b} developed a statistical model (i.e. the elongation/inclination method) that evaluates the deviation of the distribution of a large number ($>$100) of observed sedimentary paleomagnetic directional data with the predicted distribution given by a paleosecular variation model developed for the past 5 Myr (TK03) under the the geocentric axial dipole (GAD) hypothesis. Although the TK03 paleosecular variation model is based on data of relatively recent geomagnetic field, it has been shown to be compatible with data from the ca. 1.1 Ga lava flows of the Midcontinent Rift \cite{Tauxe2009a}. Further, \citeA{Pierce2022a} showed that the E/I method passes a field test for the 1.1 Ga Cut Face Creek Sandstone as the method successfully corrects the shallow paleomagnetic inclination of the sediments to that of the lava flows that bracket them. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation. High-resolution thermal demagnetization data on siltstone intraclasts that were eroded by fluvial processes and redeposited at two locations constrain the timing of hematite acquisition by revealing a primary component that formed prior to the erosion of the clasts within the depositional environment and a secondary component that formed following their redeposition. We then use a paleomagnetic fold test to further constrain the timing of detrital remanence acquisition in association with the structural inversion of the Midcontinent Rift. Finally, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia and its association with the development of the supercontinent Rodinia.

@@

@@

@@

@@

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1085-1050 Ma Oronto Group sediments \cite{Fairchild2017a, Swanson-Hysell2019a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1992a}). Subsequently, the rift was inverted as the ca. 1090-980 Ma Grenvillian Orogeny commenced on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}), eventually resulting in the propagation of far-field compressional stress deep into the interior of Laurentia \cite{Cannon1993a, Cannon1994a}.

During the inversion, Paleoproterozoic and Archean crust was uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). The southern part of the post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). More recent mapping of the Jacobsville Formation has also found that it can sometimes onlap the Midcontinent Rift volcanics within or adjacent to the Keweenaw Fault system around Bete Grise Bay on the Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{Tyrrell2019a, Mueller2021a}). Field observations find erosional contact and development of paleosol between the Jacobsville Formation and underlying volcanic rocks \cite{Hamblin1958a, Kalliokoski1975a}, and interpretations of seismic reflection data show that sedimentary rocks interpreted to be Jacobsville Sandstone lie unconformably on the Oronto Group \cite{Cannon1989a}. These observations are consistent with the interpretation that there was a prolonged period of sedimentation quiescence and net erosion between the end of Oronto Group deposition and onset of Jacobsville Formation deposition. Based on dates from Cannon et al 1993 (and deformation of the Freda Fm within the Montreal River monocline) paired with new dates on the Jacobsville Fm (Hodgin et al., 2022), this time period of quiescence may have lasted up to ~60 Myrs from ca. 1050 until ca. 990 Ma. The overall extent of the Jacobsville Formation has been mapped along the lake shore of the Keweenaw Peninsula from northern Michigan, USA, to the north side of Michipicoten Island and perhaps to northern Sault Ste Marie, Canada (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a}).

As far-field stresses from the Grenvillian Orogeny propagated into the interior of Laurentia, the previously thinned crust of the Midcontinent Rift was structurally inverted by crustal-scale thrusts, such as the Keweenaw Fault which extends $\sim$250km from ...refs here. Final inversion of the Keweenaw Fault during the Grenvillian Orogeny thrust Midcontinent Rift volcanics over the Jacobsville strata, which were folded in the footwall of the fault.  along much of the Keweenaw fault which extends $\sim$250 km from the tip of the Keweenaw Peninsula in northern Michigan southeastward to a termination in northeastern Wisconsin (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1996a, Cannon2001a, DeGraff2022a}). Syndepositional faulting that took place during Jacobsville deposition is inferred from ... rift volcanic clasts where eroded into the Jacobsville basin at the same time when sediments from the older basement rocks along with distal input from the Grenville Province were deposited within the Jacobsville back-bulge basin system that in general wedges toward the northwest (Fig. \ref{fig:Geologic_map}; \cite{Hedgman2014a, Hodgin2022a}. Next to the footwall of the Keweenaw fault, folding has also been recorded by strata of the upper Jacobsville Formation in contact with the volcanics \cite{Cannon1993a}. One famous remnant of that late-stage deformation is the Natural Wall ravine where sub-vertical sandstone beds are preserved (Fig. \ref{fig:Geologic_map}; \cite{Hamblin1958a}). On the shore of Bete Grise Bay near the northeastern end of Keweenaw Peninsula, the Keweenaw fault is interpreted to have accommodated Grenville compressional stresses by forming a series of segmented en echelon thrusts \cite{Tyrrell2019a}. The timing of such faulting could be coeval with the deposition of the upper Jacobsville Formation, leading to the nonconformable onlapping of the Jacobsville sediments over the volcanics in addition to the thrust faulting relationship between the two units. Ca. 985 Ma late-kinematic calcite that precipitated within the hanging wall of the Keweenaw Fault and ca. 992 Ma maximum deposition age from detrital zircons of the Jacobsville Formation in the footwall constrain the timing of the final inversion of the Keweenaw Fault to be during the Rigolet phase of the Grenvillian Orogeny \cite{Hodgin2022a}.

\subsection{Sedimentology and general stratigraphy}

@@

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose conglomerates, sandstones, siltstones, and shales completely devoid of lava flows or cross-cutting igneous dikes although cross-cutting clastic dikes have been observed. Abundant near the bottom section in the southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Keweenaw, Marenisco, and Pelton Creek thrust faults is interpreted to represent syn-depositional development of local relief \cite{Kalliokoski1982a, Hedgman2014a}). It is in general more quartz-rich than the Oronto Group sediments \cite{Hamblin1958a}.

Close to the Keweenaw Fault in central Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}), the Jacobsville Formation consists dominantly of quartz-rich, cross-bedded, medium-grained sandstone, with locally abundant clast-supported pebble to cobble conglomerate, and interbeds of red, hematite-bearing, micaceous shaly siltstone to fine-grained sandstone (Hungarian Falls, Hammel Creek, Natural Wall; Fig. \ref{fig:strat_column}). In this region, the conglomerate contains abundant volcanic clasts likely derived from uplifted Midcontinent Rift volcanics \cite{Irving1885a, Brojanigo1984a}. This clast-supported lithofacies can include siltstone rip-up clasts or channelized lenses of siltstones. Occurrence of red siltstone to sandstone beds are also commonly interbedded with sandstone and conglomerate \cite<e.g.>{Hamblin1958a, Irving1885a}. They typically appear brick-red to dark-red, consisting of find-grained arenite to arkosic sandstone and siltstone sometimes with minor grey reduction spots (Fig. \ref{fig:Field_photo}). Rounded detrital mica grains up to $\sim$5mm in size can be found commonly within very fine, fissile siltstone and shaly strata (Fig. \ref{fig:Field_photo}).

The exact stratigraphic association between sections of the Jacobsville Formation in the central Keweenaw Peninsula and those exposed at Agate Falls and Baby Snake Creek is difficult to constrain as the thickness of the Jacobsville strata can vary significantly through the peninsula \cite{Hamblin1958a, Kalliokoski1982a} and are not fully exposed at or in between the studied localities. As a result, various interpretations could exist for the relative stratigraphy at different localities. Based on the relative lack of conglomerate facies at Agate Falls and at Baby Snake Creek, and field observations of onlapping Jacobsville Formation on top of previously thrusted Midcontinent Rift volcanics, we interpret that Jacobsville Formation exposed at Agate Falls and Baby Snake Creek could be stratigraphically higher than the other three sections near central Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}, \ref{fig:strat_column}). Further discussion regarding the stratigraphy is presented in section \ref{magstrat}.

% Near the central part of Keweenaw Peninsula, we visited three adjacent Jacobsville stratigraphic sections at Hammel Creek, Hungarian Falls, and Natural Wall ravine (Fig. \ref{fig:Geologic_map}, \ref{fig:strat_column}). The exposed Jacobsville Formation share similarities in that they are all close to the Keweenaw Fault where the formation is steeply tilted in the footwall against the folded Midcontinent Rift volcanics in the hanging wall. They are also of similar thickness and are characterized by the interbedded layers of pebble conglomerate, massive cross-bedded sandstone and minor amount of siltstone to fine-grained sandstone (Fig. \ref{fig:strat_column}).

% The fact that in the Jacobsville has conglomerate and sandstone and siltstone facies in them supports sediment input from the eroded Oronto Group. This indicate the older boundary of the Jacobsville deposition might be closer to the Rigolet Phase than the Ottawan phase and the duration of Jacobsville sedimentation may be within a relatively constrained period of time.

% tyrrell 2019 fig. 10 recorded onlapping of the Jacobsville Formation on to the Portage Lake Volcanics which appears to have been faulted prior to the deposition of Jacobsville near Bete Grise Bay.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=0.9\textwidth]{Jacobsville_Sections_v3.pdf}
\caption{Tentative sedimentary stratigraphy and magnetostratigraphy for the Jacobsville Formation in northern Michigan, USA. Stratigraphic locations of paleomagnetic samples are shown in context of Jacobsville lithofacies, sedimentary structures, as well as detrital zircon maximum deposition ages developed by \citeA{Hodgin2022a}. }
\label{fig:strat_column}
\end{figure*}
